Reproducible Research - Peer Assignment 2: Severity Analysis of Storms in the USA (1966-2010)
========================================================

## Synopsis

**Still have to write something interesting here. Probably on the lines of what is NOAA, what kinds of events are there, what analysis was made and what kinds of conclusions.

## Introduction

## Data Processing

### Loading the data
The data is presented as a single CSV, compacted into a bzip format, which can be directly read
throught the `read.csv` command into a dataframe.
```{r cache=TRUE}
stormData = read.csv("Data/StormData.csv.bz2", stringsAsFactors=FALSE)
```

The resulting dataframe has the following structure: `r print(str(stormData))`

### Formatting the data

First, we need to generate a usable and consistent exponent to damage property. Values of "", -, ? and + have been all considered to be 0:
```{r cache=TRUE}
stormData$DMGEXP2[stormData$PROPDMGEXP %in% c("", "-", "?", "+", "0")]=0
stormData$DMGEXP2[stormData$PROPDMGEXP == "1"]=1
stormData$DMGEXP2[stormData$PROPDMGEXP %in% c("h", "H", "2")]=2
stormData$DMGEXP2[stormData$PROPDMGEXP %in% c("3", "K")]=3
stormData$DMGEXP2[stormData$PROPDMGEXP == "4"]=4
stormData$DMGEXP2[stormData$PROPDMGEXP == "5"]=5
stormData$DMGEXP2[stormData$PROPDMGEXP %in% c("M", "m", "6")]=6
stormData$DMGEXP2[stormData$PROPDMGEXP == "7"]=7
stormData$DMGEXP2[stormData$PROPDMGEXP == "8"]=8
stormData$DMGEXP2[stormData$PROPDMGEXP == "B"]=9
stormData$CDMGEXP2[stormData$CROPDMGEXP %in% c("", "?", "0")]=0
stormData$CDMGEXP2[stormData$CROPDMGEXP %in% c("M", "m")]=6
stormData$CDMGEXP2[stormData$CROPDMGEXP %in% c("K", "k")]=3
stormData$CDMGEXP2[stormData$CROPDMGEXP == "B"]=9
stormData$CDMGEXP2[stormData$CROPDMGEXP == "2"]=2
```

Having these numbers allows us to calculate which events generate the most impact. The following cutoff rules will be applied:
- Event types will be ordered in descending order by the %of total impact they generate on the entire database;
- Fatalities and injury events will be treated independently and grouped on a single data frame for health impacts
- for health impact, the events which generate 90% of impact on either fatalities or injuries will be considered;
- Property and crop damage will be treated independently and grouped on a single dataframe for economic impact;
- For economic impact, the events which generate 90% of impact on either property or damage will be considered.
```{r}
require(plyr)
quantifEvs = data.frame(event=stormData$EVTYPE, fatalities=stormData$FATALITIES,
               injuries=stormData$INJURIES, 
               propDmg = stormData$PROPDMG*10^stormData$DMGEXP2,
               cropDmg= stormData$CROPDMG*10^stormData$CDMGEXP2)
quantifEvs$event=as.character(quantifEvs$event)
sumEvs=ddply(quantifEvs, "event", summarize, 
        sumFatal=sum(fatalities), cntFatal=sum(fatalities>0), 
        sumInj=sum(injuries), cntInj=sum(injuries>0),
        sumPropDmg = sum(propDmg), cntPropDmg=sum(propDmg>0),
        sumCropDmg = sum(cropDmg), cntCropDmg=sum(cropDmg>0))
totFatal = sum(sumEvs$sumFatal)
totInj = sum(sumEvs$sumInj)
totPropDmg = sum(sumEvs$sumPropDmg)
totCropDmg = sum(sumEvs$sumCropDmg)
topFatal = order(sumEvs$sumFatal, decreasing=TRUE)
topInj = order(sumEvs$sumInj, decreasing=TRUE)
topPropDmg = order(sumEvs$sumPropDmg, decreasing=TRUE)
topCropDmg = order(sumEvs$sumCropDmg, decreasing=TRUE)
print(sprintf("The top %d event types with fatalities generate %0.1f%% of fatalities.", 22,
              sum(sumEvs$sumFatal[topFatal[1:22]])/totFatal*100))
fatalCutoff = sumEvs$event[topFatal[1:22]]
print(sprintf("The top %d event types with injuries generate %0.1f%% of injuries.", 11,
              sum(sumEvs$sumInj[topInj[1:11]])/totInj*100))
injCutoff = sumEvs$event[topInj[1:11]]
print(sprintf("The top %d event types with property damage generate %0.1f%% of losses.", 12,
              sum(sumEvs$sumPropDmg[topPropDmg[1:12]])/totPropDmg*100))
propDmgCutoff = sumEvs$event[topPropDmg[1:12]]
print(sprintf("The top %d event types with crop damage generate %0.1f%% of losses.", 14,
              sum(sumEvs$sumCropDmg[topCropDmg[1:14]])/totCropDmg*100))
cropDmgCutoff = sumEvs$event[topCropDmg[1:14]]

healthImpEvs = subset(stormData, EVTYPE %in% fatalCutoff | EVTYPE %in% injCutoff)
econImpEvs = subset(stormData, EVTYPE %in% propDmgCutoff | EVTYPE %in% cropDmgCutoff)
print("The list of relevant event types then is limited the the following:")
unique(c(healthImpEvs$EVTYPE, econImpEvs$EVTYPE))
```

The list of relevant events clearly has some quality problems, namely different spellings for the same event type. The following corrections are needed:
- Replace"EXTREME COLD" by "EXTREME COLD/WIND CHILL"
- Replace "EXTREME HEAT" by "EXCESSIVE HEAT"
- Replace "RIVER FLOOD" by "FLOOD"
- Replace "HEAT WAVE" by "HEAT"
- Replace "HURRICANE" by "HURRICANE/TYPHOON"
- Replace "RIP CURRENTS" by "RIP CURRENT"
- Replace "TSTM WIND" by "THUNDERSTORM WIND"
```{r}
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "EXTREME COLD"] = "EXTREME COLD/WIND CHILL"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "EXTREME HEAT"] = "EXCESSIVE HEAT"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "RIVER FLOOD"] = "FLOOD"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "HEAT WAVE"] = "HEAT"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "HURRICANE"] = "HURRICANE/TYPHOON"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "RIP CURRENTS"] = "RIP CURRENT"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "TSTM WIND"] = "THUNDERSTORM WIND"

healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "EXTREME COLD"] = "EXTREME COLD/WIND CHILL"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "EXTREME HEAT"] = "EXCESSIVE HEAT"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "RIVER FLOOD"] = "FLOOD"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "HEAT WAVE"] = "HEAT"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "HURRICANE"] = "HURRICANE/TYPHOON"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "RIP CURRENTS"] = "RIP CURRENT"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "TSTM WIND"] = "THUNDERSTORM WIND"
```

## Analysing the data
The following plot shows the victim counts for the event types that account for 
at least 90% of health impacts due to climatic events. Due to the large disparity 
of counts between different events, a logaritmic scale was used.
```{r}
quantifEvs = data.frame(event=healthImpEvs$EVTYPE, fatalities=healthImpEvs$FATALITIES,
               injuries=healthImpEvs$INJURIES)
quantifEvs$event=as.character(quantifEvs$event)
sumEvs=ddply(quantifEvs, "event", summarize, 
        sumFatal=sum(fatalities), cntFatal=sum(fatalities>0), 
        sumInj=sum(injuries), cntInj=sum(injuries>0))
y = rbind(data.frame(event=sumEvs$event, statType="Fatalities", statVal = sumEvs$sumFatal),
          data.frame(event=sumEvs$event, statType="Injuries", statVal = sumEvs$sumInj))
qplot(data=y, x=reorder(event, statVal, sum), y=statVal, fill=statType) +
    geom_bar(stat="identity", position="dodge")+theme(axis.text.x = element_text(angle = 90))+
    xlab("Event Type")+ylab("Victim count (log)")+
    scale_y_log10(breaks=10^seq.int(1, 5, 1), labels=comma)+
    ggtitle("Health impacts due to climatic events")
```

**Create another plot to show economic impact**
The following plot shows the victim counts for the event types that account for 
at least 90% of health impacts due to climatic events. Due to the large disparity 
of counts between different events, a logaritmic scale was used.
```{r}
quantifEvs = data.frame(event=econImpEvs$EVTYPE,
                        propDmg=econImpEvs$PROPDMG*10^econImpEvs$DMGEXP2,
                        cropDmg=econImpEvs$CROPDMG*10^econImpEvs$CDMGEXP2)
quantifEvs$event=as.character(quantifEvs$event)
sumEvs=ddply(quantifEvs, "event", summarize, 
        sumPropDmg=sum(propDmg), cntpropDmg=sum(propDmg>0), 
        sumCropDmg=sum(cropDmg), cntCropDmg=sum(cropDmg>0))
y = rbind(data.frame(event=sumEvs$event, statType="Property", statVal = sumEvs$sumPropDmg),
          data.frame(event=sumEvs$event, statType="Crop", statVal = sumEvs$sumCropDmg))
qplot(data=y, x=reorder(event, statVal, sum), y=statVal, fill=statType) +
    geom_bar(stat="identity", position="dodge")+theme(axis.text.x = element_text(angle = 90))+
    xlab("Event Type")+ylab("Damage (US$)")+
    scale_y_log10(breaks=10^seq.int(1, 5, 1), labels=comma)
```

## Results


.
