Severity Analysis of Climatic Events in the USA (1950-2011)
========================================================

## Synopsis

This study analyses data from the National Oceanic and Atmospheric Administration to evaluate the impacts of climatic events on human life and property. Data was taken from 1950 to 2011, formatted for consistent analysis and evaluated for the most impactful events. Impact severy was evaluated based on fatalities and injuries, with a cutoff of event types representing 90% of total reported numbers on either variable. A similar approach was made for economic impact, based on property and crop damage variables. Some processing of event types was done, in order to unify different spellings (abbreviations, plural forms, etc.) of the same weather event type. The resulting analysis shows that a small number of weather events are generally the most impactful for both population health and economy.

## Introduction

This report uses the [NOAA Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) to analyse severe weather events, concluding:

1. Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

In order to answer those questions, several steps were taken, as illustrated in the following sections:

1. The data was loaded from a compacted csv file
2. The data was formatted as required in order to be usable for the analysis
3. A Pareto of the events was made, concluding that a small number of event
types generate the most impacts.

## Data Processing

### Loading the data
The data is presented as a single CSV, compacted into a bzip format, which can be directly read
throught the `read.csv` command into a dataframe.
```{r cache=TRUE}
stormData = read.csv("Data/StormData.csv.bz2", stringsAsFactors=FALSE)
```

The resulting dataframe has the following structure:
```{r}
options(digits=4, scipen=100)
print(str(stormData))
```

### Formatting the data
First, we need to generate a usable and consistent exponent to damage property
(PROPDMGEXP and CROPDMGEXP). A quick analysis of the exponent values shows some
processing is required:
```{r}
unique(c(stormData$PROPDMGEXP, stormData$CROPDMGEXP))
```

It is clear that some translation will be required for usage in calculations.
Besides the H, K, M and B values, which are explained in the documentation for
the dataset, several other unexpected values appear. "", -, ? and + have been
all considered to be 0:
```{r cache=TRUE}
stormData$DMGEXP2[stormData$PROPDMGEXP %in% c("", "-", "?", "+", "0")]=0
stormData$DMGEXP2[stormData$PROPDMGEXP == "1"]=1
stormData$DMGEXP2[stormData$PROPDMGEXP %in% c("h", "H", "2")]=2
stormData$DMGEXP2[stormData$PROPDMGEXP %in% c("3", "K")]=3
stormData$DMGEXP2[stormData$PROPDMGEXP == "4"]=4
stormData$DMGEXP2[stormData$PROPDMGEXP == "5"]=5
stormData$DMGEXP2[stormData$PROPDMGEXP %in% c("M", "m", "6")]=6
stormData$DMGEXP2[stormData$PROPDMGEXP == "7"]=7
stormData$DMGEXP2[stormData$PROPDMGEXP == "8"]=8
stormData$DMGEXP2[stormData$PROPDMGEXP == "B"]=9
stormData$CDMGEXP2[stormData$CROPDMGEXP %in% c("", "?", "0")]=0
stormData$CDMGEXP2[stormData$CROPDMGEXP %in% c("M", "m")]=6
stormData$CDMGEXP2[stormData$CROPDMGEXP %in% c("K", "k")]=3
stormData$CDMGEXP2[stormData$CROPDMGEXP == "B"]=9
stormData$CDMGEXP2[stormData$CROPDMGEXP == "2"]=2
```

### Choosing relevant events
Still, the resulting list has `r length(unique(stormData$EVTYPE))` event types, many of them
having little or no impact. In order to focus on the most relevant ones, we will evaluate and
prioritize them based on their impact.

#### Health impacts
To find the most relevant events, we'll take the following steps:
- Event types will be ordered in descending order by the % of total impact they generate on the entire database;
- Fatalities and injury events will be treated independently and grouped on a 
single data frame for health impacts. The events which generate 90% of impact on
either of these will be considered;
```{r}
require(plyr)
quantifEvs = data.frame(event=stormData$EVTYPE, fatalities=stormData$FATALITIES,
               injuries=stormData$INJURIES, 
               propDmg = stormData$PROPDMG*10^stormData$DMGEXP2,
               cropDmg= stormData$CROPDMG*10^stormData$CDMGEXP2)
quantifEvs$event=as.character(quantifEvs$event)
sumEvs=ddply(quantifEvs, "event", summarize, 
        sumFatal=sum(fatalities), cntFatal=sum(fatalities>0), 
        sumInj=sum(injuries), cntInj=sum(injuries>0),
        sumPropDmg = sum(propDmg), cntPropDmg=sum(propDmg>0),
        sumCropDmg = sum(cropDmg), cntCropDmg=sum(cropDmg>0))
totFatal = sum(sumEvs$sumFatal)
totInj = sum(sumEvs$sumInj)
topFatal = order(sumEvs$sumFatal, decreasing=TRUE)
topInj = order(sumEvs$sumInj, decreasing=TRUE)
fatalCutoff = sumEvs$event[topFatal[1:22]]
injCutoff = sumEvs$event[topInj[1:11]]
healthImpEvs = subset(stormData, EVTYPE %in% fatalCutoff | EVTYPE %in% injCutoff)
```

With the variables above, it can be easily show that:
- the top **22** event types with fatalities generate **`r round(sum(sumEvs$sumFatal[topFatal[1:22]])/totFatal*100, 1)`%** of fatalities.
- the top **11** event types with injuries generate **`r round(sum(sumEvs$sumInj[topInj[1:11]])/totInj*100, 1)`%** of injuries.

### Economic impacts
A similar approach will yield the most relevant events related to economic impacts:
- Event types will be ordered in descending order by the % of total impact they generate on the entire database;
- Property and crop damage will be treated independently and grouped on a single
dataframe for economic impact. The events which generate 90% of impact on either
of these will be considered.
```{r}
totPropDmg = sum(sumEvs$sumPropDmg)
totCropDmg = sum(sumEvs$sumCropDmg)
topPropDmg = order(sumEvs$sumPropDmg, decreasing=TRUE)
topCropDmg = order(sumEvs$sumCropDmg, decreasing=TRUE)
propDmgCutoff = sumEvs$event[topPropDmg[1:12]]
cropDmgCutoff = sumEvs$event[topCropDmg[1:14]]
econImpEvs = subset(stormData, EVTYPE %in% propDmgCutoff | EVTYPE %in% cropDmgCutoff)
```

The variables calculated above allow us to see that:
- the top **12** event types with property damage generate **`r round(sum(sumEvs$sumPropDmg[topPropDmg[1:12]])/totPropDmg*100, 1)`%** of losses.
- the top **14** event types with crop damage generate **`r round(sum(sumEvs$sumCropDmg[topCropDmg[1:14]])/totCropDmg*100, 1)`%** of losses.

- After the previous steps, we conclude that the relevant event types are the
following:
`r sort(unique(c(healthImpEvs$EVTYPE, econImpEvs$EVTYPE)))`

The list of relevant events clearly has some quality problems, namely different spellings for the same event type. The following corrections are needed:
- Replace "EXTREME COLD" with "EXTREME COLD/WIND CHILL"
- Replace "EXTREME HEAT" with "EXCESSIVE HEAT"
- Replace "HEAT WAVE" with "HEAT"
- Replace "HURRICANE" with "HURRICANE/TYPHOON"
- Replace "RIVER FLOOD" with "FLOOD"
- Replace "RIP CURRENTS" with "RIP CURRENT"
- Replace "TSTM WIND" with "THUNDERSTORM WIND"
```{r}
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "EXTREME COLD"] = "EXTREME COLD/WIND CHILL"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "EXTREME HEAT"] = "EXCESSIVE HEAT"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "RIVER FLOOD"] = "FLOOD"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "HEAT WAVE"] = "HEAT"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "HURRICANE"] = "HURRICANE/TYPHOON"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "RIP CURRENTS"] = "RIP CURRENT"
econImpEvs$EVTYPE[econImpEvs$EVTYPE == "TSTM WIND"] = "THUNDERSTORM WIND"

healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "EXTREME COLD"] = "EXTREME COLD/WIND CHILL"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "EXTREME HEAT"] = "EXCESSIVE HEAT"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "RIVER FLOOD"] = "FLOOD"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "HEAT WAVE"] = "HEAT"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "HURRICANE"] = "HURRICANE/TYPHOON"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "RIP CURRENTS"] = "RIP CURRENT"
healthImpEvs$EVTYPE[healthImpEvs$EVTYPE == "TSTM WIND"] = "THUNDERSTORM WIND"
```

- After fixing the names, we are left with the following list of relevant events
(relevant being defined as being in the events that cause the top 90% of impacts):
`r sort(unique(c(healthImpEvs$EVTYPE, econImpEvs$EVTYPE)))`

## Results

### Health impacts
In order to plot the data, it first has to be formatted in a way that will allow
us to plot it adequately.
```{r cache=TRUE}
quantifEvs = data.frame(event=healthImpEvs$EVTYPE, fatalities=healthImpEvs$FATALITIES,
               injuries=healthImpEvs$INJURIES)
quantifEvs$event=as.character(quantifEvs$event)
sumEvs=ddply(quantifEvs, "event", summarize, 
        sumFatal=sum(fatalities), cntFatal=sum(fatalities>0), 
        sumInj=sum(injuries), cntInj=sum(injuries>0))
y = rbind(data.frame(event=sumEvs$event, statType="Fatalities", statVal = sumEvs$sumFatal),
          data.frame(event=sumEvs$event, statType="Injuries", statVal = sumEvs$sumInj))
```

The following plot shows the victim counts for the event types that account for 
at least 90% of health impacts due to climatic events. Due to the large variation 
of counts between different events, a logaritmic scale was used.
```{r fig.width=12}
require(ggplot2)
require(scales)
ggplot(data=y, aes(x=reorder(event, statVal, min), y=statVal, fill=statType)) +
    facet_grid(statType~.) +
    geom_bar(stat="identity", position="dodge") +
    theme(axis.text.x = element_text(angle = 90, hjust=1)) +
    labs(x="Event Type", y="Victim count (log)") +
    scale_y_log10(breaks=10^seq.int(0, 5, 1), labels=comma, limits=c(1, 100000)) +
    scale_fill_discrete(name="Impact\ntype") +
    ggtitle("Health impacts of climatic events")
```

The graph shows that only **`r length(unique(healthImpEvs$EVTYPE))`** event types 
cause **`r sum(sumEvs$sumFatal)`** fatalities and **`r sum(sumEvs$sumInj)`** injuries.

### Economic impacts
The following plot shows the victim counts for the event types that account for 
at least 90% of economic impacts due to climatic events. Due to the large disparity 
of counts between different events, a logaritmic scale was used. The commands 
before the plot format the data so that it can be properly plotted.
```{r fig.width=12}
quantifEvs2 = data.frame(event=econImpEvs$EVTYPE,
                        propDmg=econImpEvs$PROPDMG*10^econImpEvs$DMGEXP2,
                        cropDmg=econImpEvs$CROPDMG*10^econImpEvs$CDMGEXP2)
quantifEvs2$event=as.character(quantifEvs2$event)
sumEvs2=ddply(quantifEvs2, "event", summarize, 
        sumPropDmg=sum(propDmg), cntpropDmg=sum(propDmg>0), 
        sumCropDmg=sum(cropDmg), cntCropDmg=sum(cropDmg>0))
y = rbind(data.frame(event=sumEvs2$event, statType="Property", statVal = sumEvs2$sumPropDmg),
          data.frame(event=sumEvs2$event, statType="Crop", statVal = sumEvs2$sumCropDmg))
ggplot(data=y, aes(x=reorder(event, statVal, max), y=statVal, fill=statType)) +
    facet_grid(statType~.) +
    geom_bar(stat="identity", position="dodge") +
    theme(axis.text.x = element_text(angle = 90, hjust=1)) +
    labs(x="Event Type", y="Damage (log US$M)")+
    scale_y_log10(breaks=c(4000, 10^seq.int(4, 11, 1)), 
                  labels=c("4K", "10K", "100K", "1M", "10M", "100M",
                           "1B", "10B", "100B"),
                  limits=c(4000, 10^11), oob=squish) +
    scale_fill_discrete(name="Impact\ntype") +
    ggtitle("Economic impacts of climatic events")
```

The graph shows that only **`r length(unique(healthImpEvs$EVTYPE))`** event types 
cause **US$`r format(sum(sumEvs2$sumPropDmg),digits=2, big.mark=",")`** in property damage and **US$`r format(sum(sumEvs2$sumCropDmg),digits=2, big.mark=",")`** in crop losses.

### Final Conclusions
The previous graphs clearly define the most impactful weather events for the variables considered. The following table quantifies the health impacts in terms of fatalities and injuries:
```{r results='asis'}
require(xtable)

xt = xtable(arrange(sumEvs[c(1,2,4)], -sumFatal, sumInj), 
            caption="Major Weather Events: Health Impacts",
            display=c("d", "s", "d", "d"))
colnames(xt) <- c("Event", "Fatalities", "Injuries")
print(xt, type="html")
```

The following table quantifies the economic impacts in terms of property damage
and crop losses:
```{r results='asis'}
xt = xtable(arrange(sumEvs2[c(1,2,4)], -sumPropDmg, sumCropDmg), 
            caption="Major Weather Events: Economic Impacts",
            display=c("d", "s", "f", "f"))
colnames(xt) <- c("Event", "Prop Dmg (US$K)", "Crop Dmg (US$K)")
xt[,2] = xt[,2]/1000
xt[,3] = xt[,3]/1000
print(xt, type="html")
```

Still, this analysis considered events to be consistent and evenly distributed data across
the entire **61 year** period evaluated. Measurement technology and sensor distribution
have changed through this period. Also, governments have taken measures to reduce impacts
that change what events are most relevant at the beginning and end of periods,
and further analysis of changes in event severity over time are recommended.

It should also be considered that events have different distribution and impacts
over the country, and specific states and counties should rebuild this analysis
using a filter for their regions of interest.